---
title: "individual_analysis"
format: html
editor: visual
---

## Sample-wise Analysis

```{r}
# ── scDblFinder ────────────────────────────────────────────────────
renv::install("bioc::scDblFinder")

# ── celda (contains DecontX) ───────────────────────────────────────
renv::install("bioc::celda")

# ── Supporting Bioc dependencies ───────────────────────────────────
renv::install(c(
  "bioc::SingleCellExperiment",
  "bioc::BiocParallel",
  "bioc::scran",
  "bioc::scater"
))

# ── restfulr ───────────────────────────────────────────────────────
renv::install("bioc::restfulr")
renv::install("github::chris-mcginnis-ucsf/DoubletFinder")
```

### Load Libraries

```{r}
library(Seurat)
library(ggplot2)
library(dplyr)
library(scDblFinder)
library(celda)          # DecontX
library(SingleCellExperiment)
library(BiocParallel)
```

------------------------------------------------------------------------

### Step 1 — Load Matrices

> 

```{r}
base_path <- "SO_14534/CellRanger_Output/"

# Raw (unfiltered) matrices — NOT sample_filtered_feature_bc_matrix
data_SC006_filt <- Read10X(
  data.dir = file.path(base_path, "SC006", "sample_filtered_feature_bc_matrix")
)

data_SC008_filt <- Read10X(
  data.dir = file.path(base_path, "SC008", "sample_filtered_feature_bc_matrix")
)

# Create Seurat objects with minimal pre-filtering
# min.cells = 1, min.features = 100 keeps nearly everything including near-empty droplets
SC006 <- CreateSeuratObject(counts = data_SC006_filt, project = "SC006",
                             min.cells = 1, min.features = 100)
SC008 <- CreateSeuratObject(counts = data_SC008_filt, project = "SC008",
                             min.cells = 1, min.features = 100)

SC006$sample_id <- "SC006"
SC008$sample_id <- "SC008"

cat("SC006 raw dims:", dim(SC006), "\n")
cat("SC008 raw dims:", dim(SC008), "\n")
```

------------------------------------------------------------------------

### Step 2 — First-pass SCTransform + Clustering (Pre-QC)

> Clustering is done BEFORE any QC filtering so that scDblFinder has cluster context for its cluster-based doublet simulation mode.

```{r}
# ── SC006 ─────────────────────────────────────────────────────────
SC006 <- SCTransform(SC006, vst.flavor = "v2", verbose = TRUE)
SC006 <- RunPCA(SC006, verbose = FALSE)
ElbowPlot(SC006, ndims = 40)

# ── SC008 ─────────────────────────────────────────────────────────
SC008 <- SCTransform(SC008, vst.flavor = "v2", verbose = TRUE)
SC008 <- RunPCA(SC008, verbose = FALSE)
ElbowPlot(SC008, ndims = 40)
```

```{r}
# ── SC006 — set dims based on elbow ───────────────────────────────
SC006 <- FindNeighbors(SC006, dims = 1:30, verbose = FALSE)
SC006 <- FindClusters(SC006, resolution = 0.8, verbose = FALSE)
SC006 <- RunUMAP(SC006, dims = 1:30, verbose = FALSE)

# ── SC008 — set dims based on elbow ───────────────────────────────
SC008 <- FindNeighbors(SC008, dims = 1:30, verbose = FALSE)
SC008 <- FindClusters(SC008, resolution = 0.8, verbose = FALSE)
SC008 <- RunUMAP(SC008, dims = 1:30, verbose = FALSE)

# ── Visualize ─────────────────────────────────────────────────────
DimPlot(SC006, label = TRUE) + ggtitle("SC006 — Pre-QC First Pass")
DimPlot(SC008, label = TRUE) + ggtitle("SC008 — Pre-QC First Pass")
```

### Step 3 — Doublet Removal with scDblFinder (3 Consensus Iterations)

> Artificial doublets + real cells are embedded in a kNN graph. Doublet density per cell is estimated from artificial-doublet neighbourhood composition.

```{r}
# ── SC006 ─────────────────────────────────────────────────────────
sce_SC006 <- as.SingleCellExperiment(SC006)

doublet_calls_SC006 <- lapply(1:3, function(i) {
  set.seed(42 + i)
  res <- scDblFinder(sce_SC006, dbr.sd = 1)
  as.character(res$scDblFinder.class)  # convert factor to character
})

doublet_matrix_SC006    <- do.call(cbind, doublet_calls_SC006)
consensus_doublet_SC006 <- rowSums(doublet_matrix_SC006 == "doublet") >= 2
singlets_SC006          <- colnames(sce_SC006)[!consensus_doublet_SC006]

cat("SC006 doublets per iteration:\n")
cat("  iter 1:", sum(doublet_calls_SC006[[1]] == "doublet"), "\n")
cat("  iter 2:", sum(doublet_calls_SC006[[2]] == "doublet"), "\n")
cat("  iter 3:", sum(doublet_calls_SC006[[3]] == "doublet"), "\n")
cat("SC006 consensus doublets (>=2/3):", sum(consensus_doublet_SC006), "\n")
cat("SC006 singlets retained:         ", sum(!consensus_doublet_SC006), "\n")

# ── SC008 ─────────────────────────────────────────────────────────
sce_SC008 <- as.SingleCellExperiment(SC008)

doublet_calls_SC008 <- lapply(1:3, function(i) {
  set.seed(42 + i)
  res <- scDblFinder(sce_SC008, dbr.sd = 1)
  as.character(res$scDblFinder.class)  # convert factor to character
})

doublet_matrix_SC008    <- do.call(cbind, doublet_calls_SC008)
consensus_doublet_SC008 <- rowSums(doublet_matrix_SC008 == "doublet") >= 2
singlets_SC008          <- colnames(sce_SC008)[!consensus_doublet_SC008]

cat("SC008 doublets per iteration:\n")
cat("  iter 1:", sum(doublet_calls_SC008[[1]] == "doublet"), "\n")
cat("  iter 2:", sum(doublet_calls_SC008[[2]] == "doublet"), "\n")
cat("  iter 3:", sum(doublet_calls_SC008[[3]] == "doublet"), "\n")
cat("SC008 consensus doublets (>=2/3):", sum(consensus_doublet_SC008), "\n")
cat("SC008 singlets retained:         ", sum(!consensus_doublet_SC008), "\n")

# ── Subset to singlets ────────────────────────────────────────────
SC006_singlets <- subset(SC006, cells = singlets_SC006)
SC008_singlets <- subset(SC008, cells = singlets_SC008)

cat("SC006 after doublet removal:", ncol(SC006_singlets), "cells\n")
cat("SC008 after doublet removal:", ncol(SC008_singlets), "cells\n")
```

## DoubletFinder

```{r}
# # ── SC006 ─────────────────────────────────────────────────────────
# # Step 1 — pK identification (no ground truth)
# sweep_res_SC006 <- paramSweep(SC006, PCs = 1:30, sct = TRUE)
# sweep_stats_SC006 <- summarizeSweep(sweep_res_SC006, GT = FALSE)
# bcmvn_SC006 <- find.pK(sweep_stats_SC006)
# 
# # Step 2 — optimal pK
# optimal_pK_SC006 <- as.numeric(as.character(
#   bcmvn_SC006$pK[which.max(bcmvn_SC006$BCmetric)]
# ))
# cat("SC006 optimal pK:", optimal_pK_SC006, "\n")
# 
# # Step 3 — homotypic doublet proportion estimate
# annotations_SC006 <- SC006@meta.data$seurat_clusters
# homotypic_prop_SC006 <- modelHomotypic(annotations_SC006)
# nExp_SC006 <- round(0.0273 * ncol(SC006) * (1 - homotypic_prop_SC006))
# cat("SC006 expected doublets:", nExp_SC006, "\n")
# 
# # Step 4 — run DoubletFinder
# SC006 <- doubletFinder(SC006, PCs = 1:30, pN = 0.23,
#                               pK = optimal_pK_SC006,
#                               nExp = nExp_SC006, reuse.pANN = NULL,
#                               sct = TRUE)
# 
# # Check results
# # Column name is dynamic — find it
# df_col_SC006 <- grep("DF.classifications", colnames(SC006@meta.data), value = TRUE)
# table(SC006@meta.data[[df_col_SC006]])
```

------------------------------------------------------------------------

### Step 4 — Ambient RNA Decontamination with DecontX

> DecontX is run on doublet-filtered raw count matrices. Empty droplets in the raw matrix serve as the ambient RNA reference. Decontaminated counts are stored as a new assay ("decontXcounts") in each Seurat object.

```{r}
# Load raw matrices
data_SC006_raw <- Read10X(
  data.dir = file.path(base_path, "SC006", "sample_raw_feature_bc_matrix")
)
data_SC008_raw <- Read10X(
  data.dir = file.path(base_path, "SC008", "sample_raw_feature_bc_matrix")
)

# Create raw Seurat objects
# Recreate raw objects with minimal filter to remove zero-count barcodes
SC006_raw <- CreateSeuratObject(counts = data_SC006_raw, project = "SC006",
                                 min.cells = 0, min.features = 1)  # at least 1 gene
SC008_raw <- CreateSeuratObject(counts = data_SC008_raw, project = "SC008",
                                 min.cells = 0, min.features = 1)  # at least 1 gene

cat("SC006 raw dims:", dim(SC006_raw), "\n")
cat("SC008 raw dims:", dim(SC008_raw), "\n")

# Convert to SCE
sce_SC006_raw <- as.SingleCellExperiment(SC006_raw)
sce_SC008_raw <- as.SingleCellExperiment(SC008_raw)

# Run DecontX
sce_SC006_decontX <- decontX(sce_SC006_raw)
sce_SC008_decontX <- decontX(sce_SC008_raw)

# Subset to singlet barcodes
decontX_counts_SC006 <- round(decontXcounts(sce_SC006_decontX))
decontX_counts_SC006 <- decontX_counts_SC006[, colnames(SC006_singlets)]

decontX_counts_SC008 <- round(decontXcounts(sce_SC008_decontX))
decontX_counts_SC008 <- decontX_counts_SC008[, colnames(SC008_singlets)]

# Add to singlet Seurat objects
SC006_singlets[["decontXcounts"]] <- CreateAssayObject(counts = decontX_counts_SC006)
SC006_singlets$decontX_contamination <- sce_SC006_decontX$decontX_contamination[colnames(SC006_singlets)]

SC008_singlets[["decontXcounts"]] <- CreateAssayObject(counts = decontX_counts_SC008)
SC008_singlets$decontX_contamination <- sce_SC008_decontX$decontX_contamination[colnames(SC008_singlets)]

cat("SC006 assays:", Assays(SC006_singlets), "\n")
cat("SC008 assays:", Assays(SC008_singlets), "\n")
```

------------------------------------------------------------------------

### Step 5 — Switch Default Assay to decontXcounts, Compute QC Metrics

```{r}
# Switch default assay to decontaminated counts before QC
DefaultAssay(SC006_decontX) <- "decontXcounts"
DefaultAssay(SC008_decontX) <- "decontXcounts"

# Mitochondrial and hemoglobin percentages
SC006_decontX[["percent.mt"]] <- PercentageFeatureSet(
  SC006_decontX, pattern = "^MT-|^mt-", assay = "decontXcounts"
)
SC008_decontX[["percent.mt"]] <- PercentageFeatureSet(
  SC008_decontX, pattern = "^MT-|^mt-", assay = "decontXcounts"
)

SC006_decontX[["percent.hb"]] <- PercentageFeatureSet(
  SC006_decontX, pattern = "^HB[^(P)]", assay = "decontXcounts"
)
SC008_decontX[["percent.hb"]] <- PercentageFeatureSet(
  SC008_decontX, pattern = "^HB[^(P)]", assay = "decontXcounts"
)

# Pre-filter violin plots
VlnPlot(SC006_decontX,
        features = c("nFeature_decontXcounts", "nCount_decontXcounts",
                     "percent.mt", "percent.hb"),
        ncol = 4, pt.size = 0) +
  ggtitle("SC006 — Pre-filter QC (decontX counts)")

VlnPlot(SC008_decontX,
        features = c("nFeature_decontXcounts", "nCount_decontXcounts",
                     "percent.mt", "percent.hb"),
        ncol = 4, pt.size = 0) +
  ggtitle("SC008 — Pre-filter QC (decontX counts)")
```

------------------------------------------------------------------------

### Step 6 — Hard QC Filtering (Paper Thresholds)

> Thresholds from the paper: - Unique genes: 200 – 4,000 - UMIs: 200 – 20,000 - % mitochondrial: \< 10% - % hemoglobin: \< 5% (removes contaminating erythrocytes)

```{r}
SC006_filt <- subset(
  SC006_decontX,
  subset = nFeature_decontXcounts >= 200 & nFeature_decontXcounts <= 4000 &
           nCount_decontXcounts   >= 200 & nCount_decontXcounts   <= 20000 &
           percent.mt < 5 &
           percent.hb < 5
)

SC008_filt <- subset(
  SC008_decontX,
  subset = nFeature_decontXcounts >= 200 & nFeature_decontXcounts <= 4000 &
           nCount_decontXcounts   >= 200 & nCount_decontXcounts   <= 20000 &
           percent.mt < 10 &
           percent.hb < 5
)

cat("SC006 final cell count:", ncol(SC006_filt), "\n")
cat("SC008 final cell count:", ncol(SC008_filt), "\n")

# Post-filter violin plots
VlnPlot(SC006_filt,
        features = c("nFeature_decontXcounts", "nCount_decontXcounts",
                     "percent.mt", "percent.hb"),
        ncol = 4, pt.size = 0) +
  ggtitle("SC006 — Post-filter QC")

VlnPlot(SC008_filt,
        features = c("nFeature_decontXcounts", "nCount_decontXcounts",
                     "percent.mt", "percent.hb"),
        ncol = 4, pt.size = 0) +
  ggtitle("SC008 — Post-filter QC")
```

------------------------------------------------------------------------

### Step 7 — Final SCTransform + Dimensionality Reduction on Clean Cells

```{r}
finalize_seurat <- function(seu, dims = 1:20, resolution = 0.8) {
  seu <- SCTransform(seu, vst.flavor = "v2",
                     vars.to.regress = "percent.mt",
                     assay = "decontXcounts",
                     verbose = TRUE)
  seu <- RunPCA(seu, verbose = FALSE)
  ElbowPlot(seu, ndims = 40) |> print()
  seu <- FindNeighbors(seu, dims = dims, verbose = FALSE)
  seu <- FindClusters(seu, resolution = resolution, verbose = FALSE)
  seu <- RunUMAP(seu, dims = dims, verbose = FALSE)
  return(seu)
}

SC006_filt <- finalize_seurat(SC006_filt, dims = 1:20)
SC008_filt <- finalize_seurat(SC008_filt, dims = 1:30)

DimPlot(SC006_filt, label = TRUE) + ggtitle("SC006 — Final Clusters")
DimPlot(SC008_filt, label = TRUE) + ggtitle("SC008 — Final Clusters")
```

------------------------------------------------------------------------

### Step 8 — Save Cleaned Objects

```{r}
saveRDS(SC006_filt, file = "SC006_clean.rds")
saveRDS(SC008_filt, file = "SC008_clean.rds")

cat("Pipeline complete. Objects saved.\n")
```
